The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles

### 2022-11-09 ms/ms.Rmd
<edit ms>

---
title: |
  The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles 
type: ORIGINAL RESEARCH ARTICLE
author:
  - name: R. Alan Covey
    affil: a
    email: R.Alan.Covey@austin.utexas.edu
  - name: Robert Z. Selden, Jr.
    affil: b
    email: zselden@sfasu.edu
  - name: Astrid Runggaldier
    affil: c
    email: astrid@austin.utexas.edu
  - name: Nicole D. Payntar
    affil: a
    email: npayntar@utexas.edu
affiliation:
  - num: a
    address: |
      Department of Anthropology, The University of Texas at Austin
  - num: b
    address: |
      Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University 
  - num: c
    address: |
      Department of Art and Art History, The University of Texas at Austin; The Mesoamerica Center, The University of Texas at Austin
  - num: d
    address: |
      Department of Anthropology, The University of Texas at Austin
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  After centuries of looting along Peru’s North Coast, archaeologists acknowledge that the pottery of the Chimú Empire is one of the most collected Andean artifacts, but also one of the most poorly understood.  Much of the enduring classificatory uncertainty comes from the problematic provenance of most Chimú vessels, and the fact that the distinctive blackware identified as Chimú represents the production of workshops from across an extensive area, during periods of regional political decentralization (CE 900-1200), imperial growth (CE 1200-1450), and foreign conquest by the Inca (CE 1450-1535) and Spanish (after CE 1532) empires.  This chapter builds on previous seriations and field observations, using geometric morphometric analysis of a sample of 3D-scanned Chimú bottles from publicly held collections at the University of Texas at Austin.  Since Chimú bottles were formed in workshops using molds, variations in vessel shape can serve as indicators of variable practices.  We compare a sample of common Chimú blackware bottles with a sample of _Inca-Chimú_ vessels that carry features typical of the short Inca occupation of the North Coast.  Differences between the two vessel forms offer new lines of evidence that can guide new interpretations of the imperial history of Peru’s North Coast.
keywords: |
  Computational Archaeology; Museum Studies; Digital Humanities; Non-Western Art History; STEM; STEAM
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/mshape.compare.png}
\caption{Means shapes and comparisons for Chimu bottles and jars, as well as necks and rims. }
\label{fig:mshape}
\end{figure}

@@

Given the possibility that the selected bottles and jars could have been produced using either whole-vessel or partial molds, we focused the 3D analysis onto neck and basal morphology.  PCA results for both revealed significant differences between the two vessel shapes.  While bottle and jar necks are of similar diameters where they attach to the vessel body, bottle necks flare outward at their orifice, whereas jars maintain a vertical or slightly narrowing spout.  Analysis of basal shape had comparable results:  there was no significant difference in the sizes of bottle and jar bases, but the basal shape is significantly different between the two vessel types.  Jar bases are significantly more rounded than the flat bottle bases.  The analysis of these two vessel elements indicates that the morphological distinctions between bottles and jars derive from more than one vessel element.  The results do not answer the question of how molds were used to produce these vessels, although they reinforce the picture that these vessels were produced as distinct shapes.  That is, instead of hand-shaping the vessel bodies (or making them from a common press-mold) and adding distinctive mold-made necks and spouts to them, potters produced vessel bodies and necks in ways that kept the distinction between bottles and jars.

@@

Our first stage of GM analysis of Chimú blackware pottery confirms that high-resolution scanning and morphological analysis of whole vessels and key vessel elements is feasible for some of the common vessel categories found in the Chimú assemblage.  Data collection proved to be logistically straightforward, and our work with publicly-held collections provided opportunities for public education and outreach, enhancing the ethical dimensions of the new research.  The morphological analysis of whole vessels presented some challenges for working with some of the more complex shapes that appear in the Chimú assemblage, but we found bottles and jars to be amenable to analysis and abundant enough to build sizeable samples for further analysis.

The morphological comparison of bottles and jars from the AAHC revealed features that were significantly different between the two shape classes.  Although of similar sizes, the bottles and jars in our sample were distinguishable as whole vessels, and their basal and neck shapes were significantly different, as were their rim thicknesses.  These results do not resolve the broader questions about workshop practices that excavators have posed, but they suggest that the bottles and jars in our sample were produced as distinct vessels, rather than using common press molds to produce some elements (e.g., vessel bodies), or “mixing and matching” molded elements in a way that would make the two shapes harder to distinguish morphologically.  By broadening our study to include bottles and jars from the AMNH, we can consider additional characteristics of these vessels, including size differences and the use of modeled adornments (of plants, animals, and humans) on the vessels.  Such work will bring us closer to explaining key elements of Chimú ceramic variation over time, and in different parts of the north coast, offering welcome insights into the economic organization of the Chimú Empire.

>>>>

---
title: |
  The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles 
type: ORIGINAL RESEARCH ARTICLE
author:
  - name: R. Alan Covey
    affil: a
    email: R.Alan.Covey@austin.utexas.edu
  - name: Robert Z. Selden, Jr.
    affil: b
    email: zselden@sfasu.edu
  - name: Astrid Runggaldier
    affil: c
    email: astrid@austin.utexas.edu
  - name: and Nicole D. Payntar
    affil: a
    email: npayntar@utexas.edu
affiliation:
  - num: a
    address: |
      Department of Anthropology, The University of Texas at Austin
  - num: b
    address: |
      Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University 
  - num: c
    address: |
      Department of Art and Art History, The University of Texas at Austin; The Mesoamerica Center, The University of Texas at Austin
  - num: a
    address: |
      Department of Anthropology, The University of Texas at Austin
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  After centuries of looting along Peru’s North Coast, archaeologists acknowledge that the pottery of the Chimú Empire is one of the most collected Andean artifacts, but also one of the most poorly understood.  Much of the enduring classificatory uncertainty comes from the problematic provenance of most Chimú vessels, and the fact that the distinctive blackware identified as Chimú represents the production of workshops from across an extensive area, during periods of regional political decentralization (CE 900-1200), imperial growth (CE 1200-1450), and foreign conquest by the Inca (CE 1450-1535) and Spanish (after CE 1532) empires.  This study builds upon previous seriations and field observations using a geometric morphometric analysis of 3D-scanned Chimú bottles from publicly held collections at the University of Texas at Austin.  Since Chimú bottles were formed in workshops using molds, variations in vessel shape can serve as indicators of variable production practices.  We compare a sample of common Chimú blackware bottles with a sample of _Inca-Chimú_ jars that carry features typical of the short Inca occupation of the North Coast.  Differences between the two vessel forms offer new lines of evidence that can guide new interpretations of the imperial history of Peru’s North Coast.
keywords: |
  Computational Archaeology; Museum Studies; Digital Humanities; Non-Western Art History; STEM; STEAM
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

\begin{figure}\centering
\includegraphics[width=0.9\linewidth]{figs/mshape.compare.png}
\caption{Means shapes and comparisons for Chimu bottles and jars, as well as necks and rims.}
\label{fig:mshape}
\end{figure}

@@

Given the possibility that the selected bottles and jars could have been produced using either whole-vessel or partial molds, we focused the 3D analysis onto rim/neck and body/base morphology.  Results demonstrated significant differences between the two vessel shapes.  While bottle and jar necks are of similar diameters where they attach to the vessel body, bottle necks flare outward at their orifice, whereas jars maintain a vertical or slightly narrowing spout. No difference was found between body/base morphology of bottles and jars.  The results do not answer the question of how molds were used to produce these vessels, although they reinforce the picture that these vessels were produced as distinct shapes.  That is, instead of hand-shaping the vessel bodies (or making them from a common press-mold) and adding distinctive mold-made necks and spouts to them, potters produced vessel bodies and necks in ways that kept the distinction between bottles and jars.

@@

Our first stage of GM analysis of Chimú blackware pottery confirms that high-resolution scanning and morphological analysis of whole vessels and key vessel elements is feasible for some of the common vessel categories found in the Chimú assemblage.  Data collection proved to be logistically straightforward, and our work with publicly-held collections provided opportunities for public education and outreach, enhancing the ethical dimensions of the new research.  The morphological analysis of whole vessels presented some challenges for working with some of the more complex shapes that appear in the Chimú assemblage, but bottles and jars were found to be abundant and of comparable sample size.

The morphological comparison of bottles and jars from the AAHC revealed features that were significantly different between the two shape classes.  Although of similar sizes, the bottles and jars in our sample were distinguishable as whole vessels, and their basal and neck shapes were significantly different, as were their rim thicknesses.  These results do not resolve the broader questions about workshop practices that excavators have posed, but they do suggest that the bottles and jars in our sample were produced as distinct vessels, rather than using common press molds to produce some elements (e.g., vessel bodies), or _mixing and matching_ molded elements in a way that would make the two shapes harder to distinguish morphologically.  By broadening our study to include additional bottles and jars from other collections, we can begin to consider additional morphological characteristics of these vessels, including size differences and the use of modeled adornments (of plants, animals, and humans) on the vessels.  Such work will bring us closer to explaining key elements of Chimú ceramic variation over time, and in different parts of the north coast, offering welcome insights into the economic organization of the Chimú Empire.

### 2022-11-09 ms/ms.tex
<edit ms>

\author{\name{R. Alan Covey$^{a}$, Robert Z. Selden, Jr.$^{b}$, Astrid
Runggaldier$^{c}$, Nicole D. Payntar$^{a}$}
\affil{$^{a}$Department of Anthropology, The University of Texas at
Austin; $^{b}$Heritage Research Center, Stephen F. Austin State
University; Department of Biology, Stephen F. Austin State University;
and Cultural Heritage Department, Jean Monnet
University; $^{c}$Department of Art and Art History, The University of
Texas at Austin; The Mesoamerica Center, The University of Texas at
Austin; $^{d}$Department of Anthropology, The University of Texas at
Austin}
}

\thanks{CONTACT R. Alan
Covey. Email: \href{mailto:R.Alan.Covey@austin.utexas.edu}{\nolinkurl{R.Alan.Covey@austin.utexas.edu}}, Robert
Z. Selden,
Jr.. Email: \href{mailto:zselden@sfasu.edu}{\nolinkurl{zselden@sfasu.edu}}, Astrid
Runggaldier. Email: \href{mailto:astrid@austin.utexas.edu}{\nolinkurl{astrid@austin.utexas.edu}}, Nicole
D.
Payntar. Email: \href{mailto:npayntar@utexas.edu}{\nolinkurl{npayntar@utexas.edu}}}

@@

\begin{abstract}
After centuries of looting along Peru's North Coast, archaeologists
acknowledge that the pottery of the Chimú Empire is one of the most
collected Andean artifacts, but also one of the most poorly understood.
Much of the enduring classificatory uncertainty comes from the
problematic provenance of most Chimú vessels, and the fact that the
distinctive blackware identified as Chimú represents the production of
workshops from across an extensive area, during periods of regional
political decentralization (CE 900-1200), imperial growth (CE
1200-1450), and foreign conquest by the Inca (CE 1450-1535) and Spanish
(after CE 1532) empires. This chapter builds on previous seriations and
field observations, using geometric morphometric analysis of a sample of
3D-scanned Chimú bottles from publicly held collections at the
University of Texas at Austin. Since Chimú bottles were formed in
workshops using molds, variations in vessel shape can serve as
indicators of variable practices. We compare a sample of common Chimú
blackware bottles with a sample of \emph{Inca-Chimú} vessels that carry
features typical of the short Inca occupation of the North Coast.
Differences between the two vessel forms offer new lines of evidence
that can guide new interpretations of the imperial history of Peru's
North Coast.
\end{abstract}

@@

Our first stage of GM analysis of Chimú blackware pottery confirms that
high-resolution scanning and morphological analysis of whole vessels and
key vessel elements is feasible for some of the common vessel categories
found in the Chimú assemblage. Data collection proved to be logistically
straightforward, and our work with publicly-held collections provided
opportunities for public education and outreach, enhancing the ethical
dimensions of the new research. The morphological analysis of whole
vessels presented some challenges for working with some of the more
complex shapes that appear in the Chimú assemblage, but we found bottles
and jars to be amenable to analysis and abundant enough to build
sizeable samples for further analysis.

The morphological comparison of bottles and jars from the AAHC revealed
features that were significantly different between the two shape
classes. Although of similar sizes, the bottles and jars in our sample
were distinguishable as whole vessels, and their basal and neck shapes
were significantly different, as were their rim thicknesses. These
results do not resolve the broader questions about workshop practices
that excavators have posed, but they suggest that the bottles and jars
in our sample were produced as distinct vessels, rather than using
common press molds to produce some elements (e.g., vessel bodies), or
``mixing and matching'' molded elements in a way that would make the two
shapes harder to distinguish morphologically. By broadening our study to
include bottles and jars from the AMNH, we can consider additional
characteristics of these vessels, including size differences and the use
of modeled adornments (of plants, animals, and humans) on the vessels.
Such work will bring us closer to explaining key elements of Chimú
ceramic variation over time, and in different parts of the north coast,
offering welcome insights into the economic organization of the Chimú
Empire.

>>>>

\author{\name{R. Alan Covey$^{a}$, Robert Z. Selden, Jr.$^{b}$, Astrid
Runggaldier$^{c}$, and Nicole D. Payntar$^{a}$}
\affil{$^{a}$Department of Anthropology, The University of Texas at
Austin; $^{b}$Heritage Research Center, Stephen F. Austin State
University; Department of Biology, Stephen F. Austin State University;
and Cultural Heritage Department, Jean Monnet
University; $^{c}$Department of Art and Art History, The University of
Texas at Austin; The Mesoamerica Center, The University of Texas at
Austin; $^{a}$Department of Anthropology, The University of Texas at
Austin}
}

\thanks{CONTACT R. Alan
Covey. Email: \href{mailto:R.Alan.Covey@austin.utexas.edu}{\nolinkurl{R.Alan.Covey@austin.utexas.edu}}, Robert
Z. Selden,
Jr.. Email: \href{mailto:zselden@sfasu.edu}{\nolinkurl{zselden@sfasu.edu}}, Astrid
Runggaldier. Email: \href{mailto:astrid@austin.utexas.edu}{\nolinkurl{astrid@austin.utexas.edu}}, and
Nicole D.
Payntar. Email: \href{mailto:npayntar@utexas.edu}{\nolinkurl{npayntar@utexas.edu}}}

@@

\begin{abstract}
After centuries of looting along Peru's North Coast, archaeologists
acknowledge that the pottery of the Chimú Empire is one of the most
collected Andean artifacts, but also one of the most poorly understood.
Much of the enduring classificatory uncertainty comes from the
problematic provenance of most Chimú vessels, and the fact that the
distinctive blackware identified as Chimú represents the production of
workshops from across an extensive area, during periods of regional
political decentralization (CE 900-1200), imperial growth (CE
1200-1450), and foreign conquest by the Inca (CE 1450-1535) and Spanish
(after CE 1532) empires. This study builds upon previous seriations and
field observations using a geometric morphometric analysis of 3D-scanned
Chimú bottles from publicly held collections at the University of Texas
at Austin. Since Chimú bottles were formed in workshops using molds,
variations in vessel shape can serve as indicators of variable
production practices. We compare a sample of common Chimú blackware
bottles with a sample of \emph{Inca-Chimú} jars that carry features
typical of the short Inca occupation of the North Coast. Differences
between the two vessel forms offer new lines of evidence that can guide
new interpretations of the imperial history of Peru's North Coast.
\end{abstract}

@@

Our first stage of GM analysis of Chimú blackware pottery confirms that
high-resolution scanning and morphological analysis of whole vessels and
key vessel elements is feasible for some of the common vessel categories
found in the Chimú assemblage. Data collection proved to be logistically
straightforward, and our work with publicly-held collections provided
opportunities for public education and outreach, enhancing the ethical
dimensions of the new research. The morphological analysis of whole
vessels presented some challenges for working with some of the more
complex shapes that appear in the Chimú assemblage, but bottles and jars
were found to be abundant and of comparable sample size.

The morphological comparison of bottles and jars from the AAHC revealed
features that were significantly different between the two shape
classes. Although of similar sizes, the bottles and jars in our sample
were distinguishable as whole vessels, and their basal and neck shapes
were significantly different, as were their rim thicknesses. These
results do not resolve the broader questions about workshop practices
that excavators have posed, but they do suggest that the bottles and
jars in our sample were produced as distinct vessels, rather than using
common press molds to produce some elements (e.g., vessel bodies), or
\emph{mixing and matching} molded elements in a way that would make the
two shapes harder to distinguish morphologically. By broadening our
study to include additional bottles and jars from other collections, we
can begin to consider additional morphological characteristics of these
vessels, including size differences and the use of modeled adornments
(of plants, animals, and humans) on the vessels. Such work will bring us
closer to explaining key elements of Chimú ceramic variation over time,
and in different parts of the north coast, offering welcome insights
into the economic organization of the Chimú Empire.
